Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-07-11 analysis/paper/paper.qmd
updates of conclusion

over 1522 spheroids have been identified across more than 23 archaeological sites or localities in China. However, that many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited. Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each. By contrast, three sites—Xujiayao, Dingcun, and Jinsitai—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; \@SXA2014Dingcun]. Notably, the Xujiayao site alone has yielded over 1,000 specimens [@Wang2016Houjiayao].

@@

# Acknowledgements

>>>>

Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in China. However, that many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited. Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each. By contrast, three sites—Xujiayao, Dingcun, and Jinsitai—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun]. Notably, the Xujiayao site alone has yielded over 1,000 specimens [@Wang2016Houjiayao].

@@

This study provides one of the earliest systematic analyses focused on spheroids from northern China. By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the QSY spheroids. Our results demonstrate that these spheroids were produced through a programmed reduction strategy aimed at achieving standardized spherical forms, primarily using hard-hammer direct percussion. The evidence of shaping strategies on QSY spheroids confirms that they were intentionally made tools with conceptual template rather than exhausted cores. This distinguishes them from polyhedrons or multi-facial cores in their terminal reduction stages. Traditionally, lithic technology in northern China prior to the Late Pleistocene has been considered stagnant and generally classified as simple core-and-flake industries. However, the QSY assemblage from \~429ka clearly reveals a higher level of technological sophistication.

This paper also presents one of the first systematic overviews of the spatial and temporal distribution and technological variability of spheroids within China. We found that Chinese spheroids display considerable diversity in raw materials, production techniques, and morphology. Some assemblages contain spheroids that may be comparable to those from QSY and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than QSY. This highlights the particular significance of the QSY spheroids. However, the current dataset remains limited, and many sites lack detailed morphological and technological studies of spheroids, hampering broader inferences about hominin technological behaviors. We therefore emphasize the need for site-specific analysis of spheroids in future research.

Our study employed multiple quantitative methods, including spherical harmonics, some of which are still relatively novel in lithic analysis. These approaches offer useful tools for in-depth spheroid research. Given the limitations of traditional typology in capturing the functional, morphological, and technological variability of spheroids across sites, we argue that multiple method approaches are essential. From the perspective of typology, spheroids have often been dismissed as poor marker of hominin behavior due to their wide spatial and temporal distribution, combined with low-cost technologies, more refined studies increasingly reveal their diversity and potential as markers of homonin technological behavior [@tittonSubspheroidsLithicAssemblage2020; @vaquero2018]. For example, distinctions between flaking-based and battering-based spheroids—along with patterns of raw material selection, production goals, and associations with specific subsistence strategies or environmental adaptations—may prove meaningful [@Cabanes2024]. Beyond the methods of morphological and technological analysis used here, future studies incorporating functional analysis and experimental replication will further enhance our understanding.

Lastly, further research should explore the technological connections between faceted spheroids and the Acheulean. As core tools shaped through flaking, spheroids and LCTs may share underlying cognitive and manual skills. Examining their similarities and differences could offer valuable insights into the pathways of technological evolution between Oldowan and Acheulean traditions.
